---
title: "Differential expression analysis"
author: Aglaia Szukala
description: Script to perform differential expression analysis using generalized linear models (GLM)
output: html_notebook
---

Load libraries
```{r}
# 
library("edgeR")
library("RUVSeq")
library("HTSFilter")
library("gplots")
library("devtools")
library("SuperExactTest")
library("VennDiagram")
library("yarrr")
require("gridExtra")
```

Load functions
```{r}
setwd("path/to/directory")
source("DEfunctions.R")
```

Set output directory
```{r}
outdir <- "path/to/directory"
```

Trimm Genes with low counts keeping only genes with mean(cpm(x)>1) and save trimmed table of counts
```{r}
data = trimDat("countsTable.txt")
write.table(data, file ="dataFileredCounts.txt", 
            sep = "\t", quote = F)
```

Make list-based EdgeR data object
```{r}
groupGLM <- factor(c(rep("1.H",3),rep("3.H",3),rep("4.H",3), # Populations
                       rep("5.H",3),rep("1.L",3),rep("3.L",3),
                       rep("4.L",3),rep("5.L",3)))
y = makeObj(data)
```

Data normalization accounting for library size
```{r}
y <- calcNormFactors(y, method = "TMM")
```

Estimate dispersion and fit GLM (might take some time)
```{r}
GLM = fitGLM(y)
```

Define model design and contrast groups
```{r}
###Make contrast groups, using H as base expression level
ModelDesign=model.matrix(~0+groupGLM)
my.contrasts=makeContrasts("1H_L"=groupGLM1.L-groupGLM1.H, #DEG at loc 1 
                           "3H_L"=groupGLM3.L-groupGLM3.H, #DEG at loc 3
                           "4H_L"=groupGLM4.L-groupGLM4.H, #DEG at loc 4
                           "5H_L"=groupGLM5.L-groupGLM5.H, #DEG at loc 5
                           "13_L"=groupGLM1.L-groupGLM3.L,
                           "14_L"=groupGLM1.L-groupGLM4.L,
                           "15_L"=groupGLM1.L-groupGLM5.L,
                           "34_L"=groupGLM3.L-groupGLM4.L,
                           "35_L"=groupGLM3.L-groupGLM5.L,
                           "45_L"=groupGLM4.L-groupGLM5.L,
                           "13_H"=groupGLM1.H-groupGLM3.H,
                           "14_H"=groupGLM1.H-groupGLM4.H,
                           "15_H"=groupGLM1.H-groupGLM5.H,
                           "34_H"=groupGLM3.H-groupGLM4.H,
                           "35_H"=groupGLM3.H-groupGLM5.H,
                           "45_H"=groupGLM4.H-groupGLM5.H,
                           levels=ModelDesign)
```

Perform likelihood ratio test (LRT) and identify differentially expressed genes
```{r}
LRT_test(my.contrasts[,"1H_L"],GLM, col="#A71E34", 1)
LRT_test(my.contrasts[,"3H_L"],GLM, col="#00798C", 3)
LRT_test(my.contrasts[,"4H_L"],GLM, col="#F59700", 4)
LRT_test(my.contrasts[,"5H_L"],GLM, col="#6E5DAC", 5)
```

```{r}
LRT_test_loc(my.contrasts[,"13_L"],GLM,"L1","L3")
LRT_test_loc(my.contrasts[,"14_L"],GLM,"L1","L4")
LRT_test_loc(my.contrasts[,"15_L"],GLM,"L1","L5")
LRT_test_loc(my.contrasts[,"34_L"],GLM,"L3","L4")
LRT_test_loc(my.contrasts[,"35_L"],GLM,"L3","L5")
LRT_test_loc(my.contrasts[,"45_L"],GLM,"L4","L5")
LRT_test_loc(my.contrasts[,"13_H"],GLM,"H1","H3")
LRT_test_loc(my.contrasts[,"14_H"],GLM,"H1","H4")
LRT_test_loc(my.contrasts[,"15_H"],GLM,"H1","H5")
LRT_test_loc(my.contrasts[,"34_H"],GLM,"H3","H4")
LRT_test_loc(my.contrasts[,"35_H"],GLM,"H3","H5")
LRT_test_loc(my.contrasts[,"45_H"],GLM,"H4","H5")
```

Make table to be used in GOplot
```{r}
# Produce results table to be used in GOplot
makeGOplotTable(my.contrasts[,"1H_L"],GLM)
makeGOplotTable(my.contrasts[,"3H_L"],GLM)
makeGOplotTable(my.contrasts[,"4H_L"],GLM)
makeGOplotTable(my.contrasts[,"5H_L"],GLM)
```

Find overlaps between DEGs in different ecotype pairs
```{r}
findOverlapsPair(my.contrasts[,"1H_L"], my.contrasts[,"3H_L"],"#A71E34","#00798C",1,3)
findOverlapsPair(my.contrasts[,"1H_L"], my.contrasts[,"4H_L"], "#A71E34","#F59700",1,4)
findOverlapsPair(my.contrasts[,"1H_L"], my.contrasts[,"5H_L"],"#A71E34","#6E5DAC",1,5)
findOverlapsPair(my.contrasts[,"3H_L"], my.contrasts[,"4H_L"],"#00798C","#F59700",3,4)
findOverlapsPair(my.contrasts[,"3H_L"], my.contrasts[,"5H_L"],"#00798C","#6E5DAC",3,5)
findOverlapsPair(my.contrasts[,"4H_L"], my.contrasts[,"5H_L"],"#F59700","#6E5DAC",4,5)
```





